---
title: "RNA-seq Analysis Report"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    embed-resources: true
    theme: cosmo
    code-fold: true
    code-summary: "Show code"
execute:
  warning: false
  message: false
params:
  workdir: "."
  samples_file: "../config/samples.tsv"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 10, fig.height = 6, dpi = 150)

library(tidyverse)
library(jsonlite)

workdir      <- params$workdir
samples_file <- params$samples_file

sample_info <- read_tsv(samples_file, col_names = c("group", "sample"),
                        show_col_types = FALSE) %>%
  column_to_rownames("sample")
groups <- unique(sample_info$group)
```

# Sample Quality Control {#qc}

```{r qc-data}
json_files <- list.files(file.path(workdir, "01cleandata"),
                         pattern = "\\.fastp\\.json$", full.names = TRUE)

qc_data <- map_dfr(json_files, function(f) {
  js <- fromJSON(f)
  s  <- js$summary
  tibble(
    sample         = str_remove(basename(f), "\\.fastp\\.json$"),
    raw_reads      = s$before_filtering$total_reads,
    clean_reads    = s$after_filtering$total_reads,
    retention      = round(clean_reads / raw_reads * 100, 2),
    q20_before     = round(s$before_filtering$q20_rate * 100, 2),
    q30_before     = round(s$before_filtering$q30_rate * 100, 2),
    q20_after      = round(s$after_filtering$q20_rate * 100, 2),
    q30_after      = round(s$after_filtering$q30_rate * 100, 2),
    gc_content     = round(s$after_filtering$gc_content * 100, 2),
    dup_rate       = round(js$duplication$rate * 100, 2)
  )
}) %>% arrange(sample)
```

## Read Filtering Summary

```{r qc-table}
qc_display <- qc_data %>%
  select(sample, raw_reads, clean_reads, retention,
         q20_before, q20_after, q30_before, q30_after,
         gc_content, dup_rate) %>%
  rename(
    Sample      = sample,
    `Raw Reads` = raw_reads,
    `Clean Reads` = clean_reads,
    `Retention %` = retention,
    `Q20 Before %` = q20_before,
    `Q20 After %`  = q20_after,
    `Q30 Before %` = q30_before,
    `Q30 After %`  = q30_after,
    `GC %`       = gc_content,
    `Dup %`      = dup_rate
  )

knitr::kable(qc_display, format = "pipe", digits = 2,
             caption = "Per-sample read filtering statistics (fastp)")
```

## Quality Score Distribution

```{r qc-plot, fig.height=7}
qc_long <- qc_data %>%
  select(sample, q20_before, q20_after, q30_before, q30_after) %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  mutate(
    base  = ifelse(str_detect(metric, "q20"), "Q20", "Q30"),
    stage = ifelse(str_detect(metric, "before"), "Before", "After"),
    stage = factor(stage, levels = c("Before", "After"))
  )

ggplot(qc_long, aes(x = sample, y = value, fill = stage)) +
  geom_col(position = "dodge", width = 0.7) +
  facet_wrap(~ base, ncol = 2) +
  scale_fill_manual(values = c("Before" = "#95a5a6", "After" = "#2c3e50")) +
  labs(x = NULL, y = "Rate (%)",
       title = "Q20/Q30 base quality before and after filtering",
       fill = "Stage") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text = element_text(face = "bold", size = 13))
```

::: {.callout-note title="QC Methods & Parameters"}

**Trimming tool:** fastp
**Pipeline command:**
```bash
fastp -i {input.r1} -I {input.r2} \
  -o {output.r1} -O {output.r2} \
  -w 4 -h {output.report} -j {output.json}
```
**Key parameters:** `-w 4` (4 threads), default quality filter enabled.

<!-- USER: Add your custom QC notes, specific parameter explanations, or any observations about sample quality here -->

:::

---

# Alignment Statistics {#alignment}

```{r align-data}
log_files <- list.files(file.path(workdir, "03aligned"),
                        pattern = "\\.hisat2\\.log$", full.names = TRUE)

parse_hisat2_log <- function(fp) {
  txt <- readLines(fp)
  sample <- str_remove(basename(fp), "\\.hisat2\\.log$")
  overall <- unique_conc <- multi_conc <- unaligned_conc <- NA_real_
  for (line in txt) {
    if (str_detect(line, "Overall alignment rate"))
      overall <- as.numeric(str_extract(line, "[0-9]+\\.[0-9]+"))
    if (str_detect(line, "Aligned concordantly or discordantly 0 time"))
      unaligned_conc <- as.numeric(str_extract(line, "[0-9]+\\.[0-9]+"))
    if (str_detect(line, "Aligned concordantly 1 time"))
      unique_conc <- as.numeric(str_extract(line, "[0-9]+\\.[0-9]+"))
    if (str_detect(line, "Aligned concordantly >1 times"))
      multi_conc <- as.numeric(str_extract(line, "[0-9]+\\.[0-9]+"))
  }
  tibble(sample, overall, unique_conc, multi_conc, unaligned_conc)
}

align_data <- map_dfr(log_files, parse_hisat2_log) %>% arrange(sample)
```

## Alignment Summary

```{r align-table}
align_display <- align_data %>%
  rename(
    Sample                = sample,
    `Overall %`           = overall,
    `Unique Concordant %` = unique_conc,
    `Multi-mapped %`      = multi_conc,
    `Unaligned %`         = unaligned_conc
  )

knitr::kable(align_display, format = "pipe", digits = 2,
             caption = "Per-sample HISAT2 alignment statistics")
```

## Alignment Rate by Sample

```{r align-plot, fig.height=7}
align_long <- align_data %>%
  select(sample, unique_conc, multi_conc, unaligned_conc) %>%
  pivot_longer(-sample, names_to = "category", values_to = "pct") %>%
  mutate(
    category = case_when(
      category == "unique_conc"    ~ "Unique Concordant",
      category == "multi_conc"     ~ "Multi-mapped",
      category == "unaligned_conc" ~ "Unaligned"
    ),
    category = factor(category, levels = c("Unaligned", "Multi-mapped", "Unique Concordant"))
  )

ggplot(align_long, aes(x = sample, y = pct, fill = category)) +
  geom_col(width = 0.65) +
  scale_fill_manual(values = c(
    "Unique Concordant" = "#27ae60",
    "Multi-mapped"      = "#f39c12",
    "Unaligned"         = "#c0392b"
  )) +
  labs(x = NULL, y = "Percentage of read pairs (%)",
       title = "HISAT2 alignment composition by sample",
       fill = "Category") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

::: {.callout-note title="Alignment Command & Parameters"}

**Aligner:** HISAT2
**Pipeline command:**
```bash
hisat2 -x genome.hisat2 -p 10 \
  -1 {input.r1} -2 {input.r2} \
  --new-summary --rna-strandness RF \
  -S {output.sam}
```
**Key parameters:**
- `-p 10`: 10 threads
- `--rna-strandness RF`: stranded paired-end (fr-firststrand)
- `--new-summary`: machine-readable summary log

<!-- USER: Add notes about alignment parameters, reference genome version, or any alignment quality observations here -->

:::

---

# Expression Matrix {#expression}

```{r expr-data}
count_matrix <- file.path(workdir, "05expression", "genes.counts.matrix")
tmm_matrix   <- file.path(workdir, "05expression", "genes.TMM.EXPR.matrix")

counts <- read_tsv(count_matrix, show_col_types = FALSE)
tmm    <- read_tsv(tmm_matrix,   show_col_types = FALSE)

sample_names <- colnames(counts)[-1]
expr_summary <- tibble(
  sample        = sample_names,
  total_counts  = map_dbl(sample_names, ~ sum(counts[[.x]], na.rm = TRUE)),
  n_expressed   = map_dbl(sample_names, ~ sum(counts[[.x]] > 0, na.rm = TRUE)),
  pct_expressed = round(n_expressed / nrow(counts) * 100, 1)
)
```

## Matrix Overview

- **Total genes:** `r nrow(counts)`
- **Samples:** `r ncol(counts) - 1`

## Per-Sample Count Summary

```{r expr-table}
expr_display <- expr_summary %>%
  rename(
    Sample              = sample,
    `Total Counts`      = total_counts,
    `Genes Detected`    = n_expressed,
    `% of All Genes`    = pct_expressed
  )

knitr::kable(expr_display, format = "pipe",
             caption = "Per-sample expression summary")
```

## Sample Correlation Heatmap

```{r corr-heatmap, fig.height=8, fig.width=9}
tmm_mat <- tmm %>%
  column_to_rownames("feature_id") %>%
  as.matrix()

# Order samples by group
sample_order <- sample_info %>%
  rownames_to_column("sample") %>%
  arrange(group, sample) %>%
  pull(sample)

sample_cor <- cor(tmm_mat[, sample_order], method = "pearson")

cor_df <- as.data.frame(as.table(sample_cor)) %>%
  rename(V1 = Var1, V2 = Var2, corr = Freq) %>%
  mutate(
    V1 = factor(V1, levels = sample_order),
    V2 = factor(V2, levels = sample_order)
  )

ggplot(cor_df, aes(x = V1, y = V2, fill = corr)) +
  geom_tile(color = "white", size = 0.5) +
  geom_text(aes(label = round(corr, 3)), size = 3) +
  scale_fill_gradient2(low = "#2980b9", mid = "white", high = "#c0392b",
                       midpoint = median(cor_df$corr), limits = c(min(cor_df$corr), 1)) +
  labs(x = NULL, y = NULL, fill = "Pearson r",
       title = "Sample correlation (TMM-normalized, ordered by group)") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## PCA Plot

```{r pca-plot, fig.height=7, fig.width=8}
tmm_log <- log2(tmm_mat + 1)
tmm_var <- apply(tmm_log, 1, var, na.rm = TRUE)
tmm_log <- tmm_log[tmm_var > 0, ]

pca_res <- prcomp(t(tmm_log), center = TRUE, scale. = TRUE)
pca_df  <- as.data.frame(pca_res$x) %>%
  rownames_to_column("sample") %>%
  left_join(
    sample_info %>% rownames_to_column("sample"),
    by = "sample"
  )

pve <- round(summary(pca_res)$importance[2, 1:2] * 100, 1)

ggplot(pca_df, aes(x = PC1, y = PC2, color = group, label = sample)) +
  geom_point(size = 3.5) +
  geom_text(aes(label = sample), size = 3.5, vjust = -1, show.legend = FALSE) +
  labs(
    x = paste0("PC1 (", pve[1], "%)"),
    y = paste0("PC2 (", pve[2], "%)"),
    title = "PCA of TMM-normalized expression",
    color = "Group"
  ) +
  theme_minimal(base_size = 13)
```

::: {.callout-note title="Expression Analysis Methods"}

**Quantification:** featureCounts (gene-level, exon features, stranded mode 2)
**Normalization:** TMM (edgeR), applied to both counts and TPM matrices
**PCA:** log2(TMM + 1) transformed, centered and scaled

<!-- USER: Add notes about expression patterns, outlier samples, batch effects, or other observations here -->

:::

---

# Differential Expression Analysis {#de}

```{r de-data}
de_dir <- file.path(workdir, "06de_analysis", "DESeq2_Result")
de_files <- list.files(de_dir, pattern = "\\.DE_results$", full.names = TRUE)

de_summary <- map_dfr(de_files, function(fp) {
  fname    <- basename(fp)
  contrast <- str_remove(fname, "genes\\.counts\\.matrix\\.")
  contrast <- str_remove(contrast, "\\.DESeq2\\.DE_results")

  res <- read_tsv(fp, show_col_types = FALSE)
  total   <- nrow(res)
  padj05  <- sum(res$padj < 0.05, na.rm = TRUE)
  up      <- sum(res$padj < 0.05 & res$log2FoldChange > 1,  na.rm = TRUE)
  down    <- sum(res$padj < 0.05 & res$log2FoldChange < -1, na.rm = TRUE)

  tibble(contrast, total_genes = total, padj_sig = padj05, up = up, down = down)
}) %>% arrange(desc(padj_sig))
```

## DEG Summary by Contrast

```{r de-table}
de_display <- de_summary %>%
  rename(
    Contrast          = contrast,
    `Genes Tested`    = total_genes,
    `padj < 0.05`     = padj_sig,
    `Up (log2FC > 1)` = up,
    `Down (log2FC < -1)` = down
  )

knitr::kable(de_display, format = "pipe",
             caption = "Differentially expressed genes per contrast (DESeq2)")
```

## DEG Counts

```{r de-barplot, fig.height=6}
de_bar <- de_summary %>%
  select(contrast, up, down) %>%
  pivot_longer(-contrast, names_to = "direction", values_to = "count") %>%
  mutate(direction = factor(direction, levels = c("up", "down")))

ggplot(de_bar, aes(x = reorder(contrast, count), y = count, fill = direction)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = ifelse(count > 0, count, "")),
            position = position_stack(vjust = 0.5), size = 4, color = "white") +
  scale_fill_manual(
    values  = c("up" = "#c0392b", "down" = "#2980b9"),
    labels  = c("up" = "Up-regulated", "down" = "Down-regulated")
  ) +
  coord_flip() +
  labs(x = NULL, y = "Number of DEGs",
       title = "DEG counts per contrast (padj < 0.05 & |log2FC| > 1)",
       fill = "Direction") +
  theme_minimal(base_size = 13)
```

## Volcano Plots {#volcano}

::: {.panel-tabset}

```{r, results='asis'}
volcano_theme <- theme_minimal(base_size = 13) +
  theme(legend.position = "top")

for (i in seq_len(nrow(de_summary))) {
  contrast_name <- de_summary$contrast[i]
  cat(sprintf("### %s\n\n", contrast_name))

  de_file <- file.path(de_dir,
    paste0("genes.counts.matrix.", contrast_name, ".DESeq2.DE_results"))
  de_res <- read_tsv(de_file, show_col_types = FALSE) %>%
    mutate(
      neg_log10_padj = pmin(-log10(padj), 50),
      significance = case_when(
        padj < 0.05 & log2FoldChange > 1   ~ "Up",
        padj < 0.05 & log2FoldChange < -1  ~ "Down",
        TRUE                               ~ "NS"
      )
    )

  n_up   <- sum(de_res$significance == "Up")
  n_down <- sum(de_res$significance == "Down")

  p <- ggplot(de_res, aes(x = log2FoldChange, y = neg_log10_padj,
                          color = significance)) +
    geom_point(alpha = 0.5, size = 1) +
    geom_vline(xintercept = c(-1, 1), linetype = "dashed", alpha = 0.3) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", alpha = 0.3) +
    scale_color_manual(
      values = c("Up" = "#c0392b", "Down" = "#2980b9", "NS" = "#bdc3c7")
    ) +
    labs(
      x        = "log2 Fold Change",
      y        = "-log10(adjusted p-value)",
      title    = contrast_name,
      subtitle = paste0("Up: ", n_up, "  |  Down: ", n_down,
                        "  (padj < 0.05, |log2FC| > 1)"),
      color    = NULL
    ) +
    volcano_theme

  print(p)
  cat("\n\n")
}
```

:::

::: {.callout-note title="DE Analysis Methods"}

**Method:** DESeq2 (run via Trinity `run_DE_analysis.pl`)
**Significance thresholds:** padj < 0.05, |log2FC| > 1
**Contrasts file:** `config/de_contrast.tsv`

<!-- USER: Add notes about DE results interpretation, specific genes of interest, or custom thresholds here -->

:::

---

# Software Versions {#software}

```{r software-info, echo=FALSE, results='asis'}
cat("## Command-line Tools\n\n")

get_cli_version <- function(cmd, version_flag = "--version") {
  tryCatch({
    out <- system(paste(cmd, version_flag, "2>&1"), intern = TRUE)
    out <- out[nchar(trimws(out)) > 0]
    if (length(out) > 0) trimws(out[1]) else "not found"
  }, error = function(e) "not found")
}

fp_ver <- NA_character_
json_files <- list.files(file.path(workdir, "01cleandata"),
                         pattern = "\\.fastp\\.json$", full.names = TRUE)
if (length(json_files) > 0) {
  fp_ver <- tryCatch(fromJSON(json_files[1])$summary$fastp_version, error = function(e) NA)
}

cli_versions <- tibble(
  Tool = c("fastp", "HISAT2", "samtools", "Quarto"),
  Version = c(
    ifelse(is.na(fp_ver), "unknown", fp_ver),
    get_cli_version("hisat2"),
    get_cli_version("samtools"),
    get_cli_version("quarto")
  )
)
knitr::kable(cli_versions, format = "pipe",
             caption = "Command-line tool versions")

cat("\n\n## R Environment\n\n")

r_info <- tibble(
  Component = c("R", "Platform"),
  Detail = c(R.version$version.string, R.version$platform)
)
knitr::kable(r_info, format = "pipe")

cat("\n\n## R Packages\n\n")

pkg_list <- c("tidyverse", "ggplot2", "dplyr", "readr", "tidyr", "jsonlite",
              "DESeq2", "edgeR", "Rsubread", "limma")

pkgs <- tibble(Package = character(0), Version = character(0))

for (p in pkg_list) {
  ver <- tryCatch(as.character(packageVersion(p)), error = function(e) "not installed")
  pkgs <- bind_rows(pkgs, tibble(Package = p, Version = ver))
}

other_pkgs <- setdiff(names(sessionInfo()$otherPkgs), pkg_list)
for (p in other_pkgs) {
  ver <- tryCatch(as.character(packageVersion(p)), error = function(e) NA_character_)
  if (!is.na(ver)) {
    pkgs <- bind_rows(pkgs, tibble(Package = p, Version = ver))
  }
}

knitr::kable(pkgs, format = "pipe",
             caption = "R package versions used in this analysis")
```
